Open archaeology, open source?

### 2024-05-23 analysis/_01-introduction.qmd
Revisions based on editorial comments.

A 2012 special issue of *World Archaeology* marked the coming of age of 'open archaeology',
a new, digitally-oriented archaeology "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity" [@beck2012 480-481],
situated within the wider the open science movement [@lake2012 and @beck2012].
In the same issue, @ducke2012 specifically called attention to software, calling on archaeologists to more actively engage in open source practices – making software with source code available for anyone to freely access, modify or reuse.
Open source, though originating outside academy, emerged as an important component and complement of open science, sharing its vision for a communal, self-correcting and transparent mode of knowledge production.
Where before archaeologists had relied almost exclusively on "hand-me-down" software re-purposed from other fields [@scollar1999], open archaeology envisaged a community of archaeological research software engineers oriented around open source development practices and tools.

More than decade on, we are in a position to ask whether this hopeful early rhetoric of open archaeology has been borne out in actually-existing open source research software engineering in the field. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
In other words, is there more to 'open archaeology' than just uploading text files to the internet?

@@

We use quantitative analysis to consider how archaeological software development may be benefiting from, or missing out on, the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency.

>>>>

A 2012 special issue of *World Archaeology* marked the coming of age of 'open archaeology',
a new, digitally orientated archaeology "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity" [@beck2012 480-481],
situated within the wider the open science movement [@lake2012 and @beck2012].
In the same issue, @ducke2012 specifically drew attention to software -- the programs and other operating information used by a computer to analyse archaeological information -- calling on archaeologists to more actively engage in open-source practices – making software with source code available for anyone to freely access, modify or reuse.
Open source, though originating outside academy, emerged as an important component and complement of open science, sharing its vision for a communal, self-correcting and transparent mode of knowledge production.
Where before archaeologists had relied almost exclusively on "hand-me-down" software re-purposed from other fields [@scollar1999], open archaeology envisaged a community of archaeological research software engineers oriented around open-source development practices and tools.

More than decade on, we are in a position to ask whether this hopeful early rhetoric of open archaeology has been borne out in actually existing open-source research software engineering in the field. 
Does academic open source actually make research processes more transparent and improve research outcomes? 
Is it actually boosting efficiency by establishing a common store of knowledge and productive code? 
Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass?
In other words, is there more to 'open archaeology' than just uploading text files to the internet?

@@

We use quantitative analysis to consider how archaeological software development may be benefiting from, or missing out on, the affordances that open-source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency.
As such, our work examines whether archaeologists are harnessing the collaborative potential that the open science movement ascribes to the use of open source software and resources.

### 2024-05-23 analysis/_02-background.qmd
Revisions based on editorial comments.

The idea that scientists should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals [@mirowski2018]. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by information and communication technologies that host files, document processes, facilitate co-working opportunities, and perform automated processes (e.g. digital repositories, reproducible notebooks, automated evaluation of research findings, systems for ensuring stable references to consistent identifiers, etc). However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science [@leonelli2023: 23-24].

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concerns over consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions.^[Though, participatory or community-based research does exist and is an exception to this generalization (see @morgan2012). It requires active effort to do well and should not be considered a passive by-product of posting one's research on the web.]
Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs [@kling2003]. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of resume-building, tenure and promotion [@mirowski2018; @dorta-gonzalez2021]. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out [@carver2022; @adema2021]. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours [@kelty2008: 271-275; @howison2013]. Scientists may archive their work in an institutional repository, at which point the work enters a stasis state that invites reference to the completed work, but which precludes any potential for the work to be updated, modified, or directly built upon. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software [@kelty2008: 278-281; @coleman2012: 116-122; @hippel2003].

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scientific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], Huggett [-@huggett2018; -@huggett2022], @sobotkova2018, @opitz2021, @haciguzeller2021, and @batist2023a demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. In this paper we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

>>>>

The idea that scientists should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals [@mirowski2018]. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by information and communication technologies that host files, document processes, facilitate co-working opportunities, and perform automated processes (e.g. digital repositories, reproducible notebooks, automated evaluation of research findings, systems for ensuring stable references to consistent identifiers, etc).

This culminates in an obsessive concern with digital workflows pertaining to legal and logistical issues; for many proponents of open science, publishing is largely considered the business of typesetting and copyright law, which could be rendered moot by using automated publishing workflows and by encouraging use of open licensing agreements (cf. Foster and Deardorff 2017; Harnad 1998). Viewed as merely technical systems, these could be resolved through technical means. However academic publishing and ownership involve social arrangements that serve to stabilise knowledge, grant authority to validated claims, and enable science to move forward (Kelty et al. 2008: 274-275). In other words, technocentric visions of publication workflows tend to ignore the fact that publication is a cultural phenomenon, whereby projects are made complete and knowledge claims are articulated, credited, and rendered accountable to the people who proposed them. However, technological systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science [@leonelli2023: 23-24].

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concerns over consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science is driven by a warrant to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions.^[Though, participatory or community-based research does exist and is an exception to this generalization (see @morgan2012). It requires active effort to do well and should not be considered a passive by-product of posting one's research on the web.]
Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs [@kling2003]. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of resume-building, tenure and promotion [@mirowski2018; @dorta-gonzalez2021]. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out [@carver2022; @adema2021]. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours [@kelty2008: 271-275; @howison2013]. Scientists may archive their work in an institutional repository, at which point the work enters a stasis state that invites reference to the completed work, but which precludes any potential for the work to be updated, modified, or directly built upon. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software [@kelty2008: 278-281; @coleman2012: 116-122; @hippel2003].

The adoption of open source development models among archaeologists is generally informed by the broader open science movement. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], Huggett [-@huggett2018; -@huggett2022], @sobotkova2018, @opitz2021, @haciguzeller2021, and @batist2023a demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research -- such as the inherent subjectivities involved in characterising finds, or the reliance on analysts’ reputation and academic pedigree to establish trust in the data they produce -- assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. In this articlle we draw attention to similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists by analysing the collaborative milieu in which archaeological software are actually being built and relating our findings to broader observations about how archaeologists tend to collaborate in practice.

### 2024-05-23 analysis/_03-data_methodology.qmd
Revisions based on editorial comments.

We compiled the dataset by browsing collaborative software development platforms, relying heavily on their social networking features.
More specifically, we update open-archaeo by manually crawling through archaeologists' profiles on these platforms, as well as on other personal, professional, and institutional websites that describe and host additional archaeological software.
We supplement this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals who identified relevant work that we initially overlooked.

@@

Open-archaeo is a relatively comprehensive list.
While our initial intention was to only list open source software, its scope has expanded to include all software created by and for archaeologists.
Apart from regular updates by its primary maintainer (ZB), it has been expanded by a wider network of contributors and has benefited from the wider range of domain specialisms this has brought.
However, open-archaeo generally lacks software written before archaeologists started using collaborative software development platforms such as GitHub, and software that is not shared on the web at all.
It also includes numerous non-software resources, as well as software developed and distributed without the use of software forges.
The dataset is also limited by the experiences of its primary maintainers.^[We welcome anyone, especially domain specialists who are familiar with the kinds of tools commonly used in their specific fields, to help fill in these gaps. Instructions for contributing to open-archaeo can be found at <https://github.com/zackbatist/open-archaeo>.]

@@

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects
oarch_forges |>
  mutate(
    repo_host = recode(
      repo_host,
      "GitHub" = "[GitHub](https://github.com/)",
      "Codeberg" = "[Codeberg](https://codeberg.org/)",
      "GitLab" = "[GitLab](https://about.gitlab.com/)",
      "BitBucket" = "[Bitbucket](https://bitbucket.org/product/)",
      "Launchpad" = "[Launchpad](https://launchpad.net/)"
    ),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  rename(Host = repo_host, `%` = p) |>
  tt() |>
  format_tt(1, markdown = TRUE)
```

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
Our analysis incorporates data on `r n_repos` repositories^[We excluded three GitHub repositories from the analysis for technical reasons. For example, https://github.com/carpentries-incubator/R-archaeology-lesson is a repository within the scope of open-archaeo, but it was forked from a pre-existing repository that is not (https://github.com/datacarpentry/R-ecology-lesson) and thus includes in its commit history irrelevant data from the parent repository.], comprising `r n_commits` commits, `r n_issues` issues/pull requests, and `r n_comments` from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

@@

That being said, we cannot account for practices that occur through offline or private channels, or forms of collaboration we do not know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves. 
Our earliest data is from 2005 and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013; @whallon1972].

These caveats notwithstanding, the open-archaeo directory and the supplemental data from the GitHub API provide a rich resource to explore the nature of collaborative software engineering in archaeology.
Here we employ exploratory data analysis [*sensu* @tukey1977] to identify and describe overall patterns visible in this rich dataset.
In @sec-open-archaeology, our focus is on examining the general state of open source archaeological software and resource development.
In @sec-collaboration, we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in @sec-network, we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, and how they create these tools in response to specific needs and use-cases, as afforded by offline social and professional connections, and within the context of an emerging community of practice.

>>>>

We compiled the dataset by browsing collaborative software development platforms, relying heavily on their social networking features.
More specifically, we update open-archaeo by manually crawling through archaeologists' profiles on these platforms, as well as on other personal, professional, and institutional websites that describe and host additional archaeological software.
We supplement this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals, who reached out via email, social media or at conferences to identify relevant work that we initially overlooked, including work that they created themselves.

@@

Open-archaeo is a relatively comprehensive list.
While our initial intention was to only list open source software, its scope has expanded to include all software created by and for archaeologists.
Apart from regular updates by its primary maintainer (ZB), it has been expanded by a wider network of contributors and has benefited from the wider range of domain specialisms this has brought.
However, open-archaeo generally lacks software written before archaeologists started using collaborative software development platforms such as GitHub, and software that is not shared on the web at all.
It also includes numerous non-software resources, as well as software developed and distributed without the use of software forges.
Open-archaeo is also limited by the experiences of its primary maintainers, which affects the dataset's overall scope and how comprehensively it covers various domains of archaeological research.^[We welcome anyone, especially domain specialists who are familiar with the kinds of tools commonly used in their specific fields, to help fill in these gaps. Instructions for contributing to open-archaeo can be found at <https://github.com/zackbatist/open-archaeo>.]

@@

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects, as per open-archaeo
oarch_forges |>
  mutate(
    repo_host = recode(
      repo_host,
      "GitHub" = "[GitHub](https://github.com/)",
      "Codeberg" = "[Codeberg](https://codeberg.org/)",
      "GitLab" = "[GitLab](https://about.gitlab.com/)",
      "BitBucket" = "[Bitbucket](https://bitbucket.org/product/)",
      "Launchpad" = "[Launchpad](https://launchpad.net/)"
    ),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  rename(Host = repo_host, `%` = p) |>
  tt() |>
  format_tt(1, markdown = TRUE)
```

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API (application programming interface).
Our analysis incorporates data on `r n_repos` repositories^[We excluded three GitHub repositories from the analysis for technical reasons. For example, https://github.com/carpentries-incubator/R-archaeology-lesson is a repository within the scope of open-archaeo, but it was forked from a pre-existing repository that is not (https://github.com/datacarpentry/R-ecology-lesson) and thus includes in its commit history irrelevant data from the parent repository.], comprising `r n_commits` commits and `r n_issues` issues/pull requests from `r n_contributors` distinct users, as well as repository metadata on programming languages used, licensing, stars and forks, and so on.

@@

That being said, we cannot account for practices that occur through offline or private channels, or forms of collaboration we do not know about. 
We did not directly observe or interview archaeological software developers, though our conclusions do draw heavily from our experience as members of that community ourselves.^[JR is an extremely prolific archaeological open source software developer, who actively builds code, teaches programming methods, and conducts code reviews. Through his maintenance of open-archaeo over the past 5+ years, ZB has honed an extensive understanding of the people who create archaeological software and the institutional ecology that supports their work.] 
Our earliest data is from 2005 and our study can say little about collaborative software development in archaeology before this point, though we know there was a significant amount of it [@ducke2013; @whallon1972].

These caveats notwithstanding, the open-archaeo directory and the supplemental data from the GitHub API provide a rich resource to explore the nature of collaborative software engineering in archaeology.
Here we employ exploratory data analysis [*sensu* @tukey1977] to identify and describe overall patterns visible in this rich dataset.
In @sec-open-archaeology, our focus is on examining the general state of open source archaeological software and resource development.
In @sec-collaboration, we refine our analysis to examine development processes, with specific focus on collaborative experiences.
Finally, in @sec-network, we apply network analysis methods to investigate the formation of broader collaborative communities.
Our analyses combine to support our objectives of understanding what kinds of software and resources archaeologists are making, how they create these tools in response to specific needs and use-cases, and how this work is situated within the context of an emerging community of practice.

### 2024-05-23 analysis/_04-open_archaeology.qmd
Revisions based on editorial comments.

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums.^[see [https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code](https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code) for a similar list of archaeology publications that include R code]
We annotated each record with categories and tags that describe what aspect of archaeological research each tool or resource was meant to address.
We also categorized all records based on how each tool or resource is meant to be accessed or used, and assigned tags based on how developers identified their projects' purpose and scope.
See @batist2023 for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

n_platform <- sum(!is.na(oarch$platform))
p_platform <- n_platform / nrow(oarch)

@@

n_datasets <- nrow(filter(oarch, tag1 == "Datasets" | tag2 == "Datasets" | 
                            tag3 == "Datasets" | tag4 == "Datasets" | 
                            tag5 == "Datasets"))
p_datasets <- n_datasets / nrow(oarch)

n_scripts <- nrow(filter(oarch, category == "Scripts"))
p_scripts <- n_scripts / nrow(oarch)
```

In our breakdown of open-archaeo by category (see @tbl-categories), we demonstrate the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language (e.g. radiocarbon calibration is implemented in R in the package 'rcarbon' or Python in the package 'iosacal') or a plugin for an application (e.g. 'ArchaeoAstroInsight' adds tools for measuring astronomical alignments to QGIS).
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  mutate(
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(platform == "Other", -n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

```{r data-platforms}
p_platform_r <- sum(oarch$platform == "R", na.rm = TRUE) / nrow(oarch)
p_platform_python <- sum(oarch$platform == "Python", na.rm = TRUE) / nrow(oarch)
p_platform_qgis <- sum(oarch$platform == "QGIS", na.rm = TRUE) / nrow(oarch)
```

`r percent(p_platform_r)` of projects are extensions to the statistical programming language R, making it the most widely-used platform by a large margin (@tbl-platforms).
Python, another programming language, is also relatively popular (`r percent(p_platform_python)`), as are plugins for the open source geographic information system QGIS (`r percent(p_platform_qgis)`).^[While Blender and QGIS plugins are written using the Python language, our intent while categorizing platforms was to get a sense of the developer ecosystems in which archaeological software engineers participate, rather than to simply gauge the popularity of different languages [@batist2023 2].]
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

p_license_gpl <- n_filter(oarch, str_detect(license, "GPL")) / n_filter(oarch, !is.na(license))
p_license_mit <- n_filter(oarch, str_detect(license, "MIT")) / n_filter(oarch, !is.na(license))
p_license_cc <- n_filter(oarch, str_detect(license, "CC")) / n_filter(oarch, !is.na(license))
```

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license (@tbl-repo-licenses).
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of open source software within non-open software, [@dusollier2007].
Interestingly, archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

>>>>

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums (collections of digital resources, including data, code, and documentation, which accompany or enhance a scientific publication; see [https://research-compendium.science](https://research-compendium.science).).^[see [https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code](https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code) for a similar list of archaeology publications that include R code]
Each record in open-archaeo is is assigned to a category based on how the tool or resource is meant to be accessed or used, and is annotated with tags that describe what aspect of archaeological research each itemtool or resource was meant to address.
Tags are ascribed based on how developers identified their projects’ purpose and scope, and each record can have multiple tags.
See @batist2023 for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

n_packages <- nrow(filter(oarch, category == "Packages and libraries"))
p_packages <- n_packages / nrow(oarch)

@@

n_scripts <- nrow(filter(oarch, category == "Scripts"))
p_scripts <- n_scripts / nrow(oarch)

n_datasets <- nrow(filter(oarch, category == "Lists and datasets"))
p_datasets <- n_datasets / nrow(oarch)

# n_platform <- sum(!is.na(oarch$platform))
# p_platform <- n_platform / nrow(oarch)

# n_datasets <- nrow(filter(oarch, tag1 == "Datasets" | tag2 == "Datasets" | 
#                             tag3 == "Datasets" | tag4 == "Datasets" | 
#                             tag5 == "Datasets"))
# p_datasets <- n_datasets / nrow(oarch)
```

From our breakdown of open-archaeo by category (see @tbl-categories), we can infer the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_packages)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language (e.g. radiocarbon calibration is implemented in R in the package 'rcarbon' or Python in the package 'iosacal') or a plugin for an application (e.g. 'ArchaeoAstroInsight' adds tools for measuring astronomical alignments to QGIS).
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

```{r tbl-platforms}
#| tbl-cap: Platforms and programming languages used by open archaeology projects
oarch |>
  drop_na(platform) |>
  count(platform) |>
  mutate(
    significant = n > 1,
    platform = if_else(significant, platform, "Other")
  ) |>
  group_by(platform) |>
  summarise(n = sum(n)) |>
  mutate(
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(platform == "Other", -n) |>
  rename_with(tools::toTitleCase) |>
  tt()
```
```{r data-platforms}
  p_platform_r <- sum(oarch$platform == "R", na.rm = TRUE) / sum(!is.na(oarch$platform))
  p_platform_python <- sum(oarch$platform == "Python", na.rm = TRUE) / sum(!is.na(oarch$platform))
  p_platform_qgis <- sum(oarch$platform == "QGIS", na.rm = TRUE) / sum(!is.na(oarch$platform))
  ```

Some `r percent(p_platform_r)` of projects are extensions to the statistical programming language R, making it the most widely-used platform by a large margin (@tbl-platforms).
Python, another programming language, is also relatively popular (`r percent(p_platform_python)`), as are plugins for the open source geographic information system QGIS (`r percent(p_platform_qgis)`).^[While Blender and QGIS plugins are written using the Python language, our intent while categorizing platforms was to get a sense of the developer ecosystems in which archaeological software engineers participate, rather than to simply gauge the popularity of different languages [@batist2023 2].]
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

p_license_gpl <- n_filter(oarch, str_detect(license, "GPL")) / sum(!is.na(oarch$license))
p_license_mit <- n_filter(oarch, str_detect(license, "MIT")) / sum(!is.na(oarch$license))
p_license_cc <- n_filter(oarch, str_detect(license, "CC")) / sum(!is.na(oarch$license))
```

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license (@tbl-repo-licenses).
Two common free software licenses account for the majority of repositories that do contain licenses: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of open source software within non-open software, [@dusollier2007].
Interestingly, archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

### 2024-05-23 analysis/_05-collaboration.qmd
Revisions based on editorial comments.

We measured the lifespan of a repository as the time between the first and latest commit, and its activity as the rate of commits.
Here therefore we refer to the *development* lifespan of a project, which is not necessary related to its use-life.
By these metrics, the lifespan and activity of repositories in open-archaeo vary greatly (@fig-lifespan).
The average project lasts `r round(mean(oarch_lifespan$lifespan))` days with `r round(mean(oarch_lifespan$commit_rate, na.rm = TRUE), 2)` commits per day.
Many projects are active for only a short period of time: about `r percent(p_lifespan_30days)` less than 30 days, `r percent(p_lifespan_90days)` less than 90 days, and `r percent(p_lifespan_365days)` less than a year.
However, the vast majority (all but `r english(n_1commit)`) do have more than one commit, suggesting that use of GitHub as a pure host for already-finished projects is not common; some degree of iteration, if not collaboration, is almost always present.
The longest-lived projects have been active for between `r english(round(lifespan_top10_range / 365)[1])` and `r english(round(lifespan_top10_range / 365)[2])` years.
The most active projects see up to `r round(max_commit_rate)` commits per day, but the majority of repositories (`r percent(p_commit_rate_1)`) receive less one commit per day.

@@

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_75)`.
This may be attributed to the steep learning curve commonly attributed to working with git.
While git can be a great way to track changes and manage distributed contributions to a common code base, it can also be unwieldy in situations when multiple users (especially those with less experience using git for collaborative purposes) are expected to contribute within short spans of time.
This meshes with our prior observations that projects exhibiting higher commit rates have less contributors.
Additionally, our analyses neglect to account for contributions that are not tracked via git or GitHub.
Those who do not code may provide creative guidance or feedback during in-person meetings, via email, or using alternative online messaging or social media platforms.
A more focused qualitative assessment of these non-coding and supportive work practices would shed more light on the totality of effort that goes into producing and maintaining open source projects.

>>>>

We measured the lifespan of a repository as the time between the first and latest commit, and its activity as the number of commits per day.
Here therefore we refer to the *development* lifespan of a project, which is not necessary related to its use-life.
By these metrics, the lifespan and activity of repositories in open-archaeo vary greatly (@fig-lifespan).
The average project lasts `r round(mean(oarch_lifespan$lifespan))` days with `r round(mean(oarch_lifespan$commit_rate, na.rm = TRUE), 2)` commits per day.
Many projects are active for only a short period of time: about `r percent(p_lifespan_30days)` less than 30 days, `r percent(p_lifespan_90days)` less than 90 days, and `r percent(p_lifespan_365days)` less than a year.
However, the vast majority (all but `r english(n_1commit)`) do have more than one commit, suggesting that use of GitHub as a pure host for already-finished projects is not common; some degree of iteration, if not collaboration, is almost always present.
The longest-lived projects have been active for between `r english(round(lifespan_top10_range / 365)[1])` and `r english(round(lifespan_top10_range / 365)[2])` years.
The most active projects see up to `r round(max_commit_rate)` commits per day, but the majority of repositories (`r percent(p_commit_rate_1)`) receive less one commit per day.

@@

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_75)`.
This may be attributed to the steep learning curve commonly attributed to working with git.
While git can be a great way to track changes and manage distributed contributions to a common code base, it can also be unwieldy in situations when multiple users (especially those with less experience using git for collaborative purposes) are expected to contribute within short spans of time.
This corresponds with our prior observations that projects exhibiting higher commit rates have less contributors.
Additionally, our analyses neglect to account for contributions that are not tracked via git or GitHub.
Those who do not code may provide creative guidance or feedback during in-person meetings, via email, or using alternative online messaging or social media platforms.
A more focused qualitative assessment of these non-coding and supportive work practices would shed more light on the totality of effort that goes into producing and maintaining open source projects.
